Technique for \earlywarning\ detection of compact binary coalescence and its implication for multimessenger astronomy

### 2011-06-02 analysis.tex
some more text
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@827 ccdd05d5-56c1-461a-bf98-85f6876523b4
more revisions
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@833 ccdd05d5-56c1-461a-bf98-85f6876523b4

In this section we describe an implementation of the \lloid\ method described
in section \ref{label{sec:method}} suitable for rapid gravitational-wave
searches for compact binary coalescence.  The \lloid\ method requires several
computations that can be done before the analysis is actually underway.  Thus
we divide the procedure into two stages 1) an offline planning stage and 2) an
online, low-latency filtering stage.  The offline stage can be done before the
analysis is started and updated asychronously, whereas the online stage must
keep up with the detector output and produce search results as rapidly as
possible.  In the next two subsections we describe what these stages entail.

@@

The planning stage begins with choosing templates that cover the space of mass
parameters with a hexagonal grid \cite{PhysRevD.76.102004} in order to satisfy
the minimum match criterion, which assures a user specified maximum loss in SNR
for signals that fall in-between the chosen templates.  Next, the templates are
subdivided into groups of neighbors called ``sub-banks'' that are appropriately
sized so that each bank can be efficiently handled by a single computer.
Dividing the mass space into smaller sub-banks also aids in the computational
cost of the singular value decomposition and is the approach considered
in~\cite{Cannon:2010p10398}.  Using our understanding of the time-frequency
evolution of the templates, we choose time slice boundaries as in \eqref{FIXME}
such that all of the templates within a sub-bank are sub-critically sampled at
progressively lower sample rates.  Next, the templates within the sub-bank are
realized as \textsc{fir} filter coefficients.  For each time slice, the
templates are downsampled to the appropriate sample rate.  Finally, the
\textsc{svd} is applied to each time slice in the sub-bank in order to produce
a set of orthogonal \textsc{fir} filters and a reconstruction matrix that maps
them back to the original templates as described in \eqref{FIXME}.  The
downsampled orthogonal \textsc{fir} filter coefficients, the reconstruction
matrix, and the time slice boundaries are all saved to disk.

@@

We have implemented a prototype of the low latency filtering stage using an
open source signal processing environment called \gstreamer\ \cite{gstreamer}.
\gstreamer\ is a vital component of many Linux systems, providing media
playback, authoring, and streaming on devices from cell phones to desktop
computers to streaming media servers.  Given the similarities of gravitational
wave detector data to audio data it is not surprising that \gstreamer\ is
useful for our purpose. \gstreamer\ also provides some useful stock signal
processing elements such as decimators, filters, and interpolators.  
We have extended the \gstreamer\ framework by developing custom elements
that are available in the \gstlal\ library~\cite{gstlal}.
Figure \ref{fig:pipeline} describes schematically how we implement the \lloid\
algorithm using \gstlal\ and \gstreamer\ components.
%
%
\begin{figure}[htbp]
	\includegraphics{figures/lloid-diagram.pdf}
	\caption{\label{fig:pipeline} Schematic of LLOID pipeline illustrating
signal flow.  Circles with arrows represent upsampling
\protect\includegraphics{figures/upsample-symbol.pdf} or downsampling
\protect\includegraphics{figures/downsample-symbol.pdf}.  Circles with plus
signs represent summing junctions
\protect\includegraphics{figures/adder-symbol.pdf}.  Squares
\protect\includegraphics{figures/fir-symbol.pdf} stand for FIR filters.  Sample
rate decreases from the top of the diagram to the bottom.}
\end{figure}

The filter pipeline drawn in figure \ref{fig:pipeline} consists of several
distinct stages and parallel branches.  We outline the stages below.

@@

First, the sample rate of the whitened detector data is reduced to successively
lower sample rates by decimation.  Decimation involves applying an antialiasing
filter to the data, and then downsampling by deleting samples.  We use a
192-tap \textsc{fir} decimator provided by {\tt Gstreamer's audioresample}
element.  The detector data is provided at every power-of-two sample rate
required by the template time sliced described in \eqref{FIXME}.  These
parallel decimated data streams are fed into parallel fir filter banks in the
next stage.

\subsubsection{Orthogonal \textsc{fir} filters}

@@

The orthogonal filter output provided by the SVD must be used in linear
combinations in order to obtain the physical filter output for the template
waveforms.  This involves a matrix multiplication.  We use a \gstlal\ element
called {\tt lal\_matrixmixer} on the output of each orthogonal filter bank in
order to produce the filter output for the original waveforms in each time
slice.  At this point the time slice filter outputs are not yet at the same
sample rate.

@@

>>>>

In this section we describe an implementation of the \lloid\ method described
in section \ref{sec:method} suitable for rapid gravitational-wave
searches for compact binary coalescence.  The \lloid\ method requires several
computations that can be completed before the analysis is underway.  Thus
we divide the procedure into two stages 1) an offline planning stage and 2) an
online, low-latency filtering stage.  The offline stage can be done before the
analysis is started and updated asychronously, whereas the online stage must
keep up with the detector output and produce search results as rapidly as
possible.  In the next two subsections we describe what these stages entail.

@@

The planning stage begins with choosing templates that cover the space of mass
parameters with a hexagonal grid~\cite{PhysRevD.76.102004} in order to satisfy
a minimum match criterion.  This assures a user-specified, maximum loss in SNR
for signals that fall in-between the chosen templates.  Typically the minimum
match is 97\% corresponding to a maximum mismatch of 3\%.  Next, the templates
are subdivided into groups of neighbors called ``sub-banks'' that are
appropriately sized so that each bank can be efficiently handled by a single
computer.  The neighbors are chosen to have similar chirp mass, which produces
subbanks with very similar waveforms.  Dividing the mass space into smaller
sub-banks aids in the computational cost of the singular value decomposition
and is the approach considered in~\cite{Cannon:2010p10398}.  Using our
understanding of the time-frequency evolution of the templates, we choose time
slice boundaries as in \eqref{FIXME} such that all of the templates within a
sub-bank are sub-critically sampled at progressively lower sample rates.  Next,
the templates within the sub-bank are realized as \fir\ filter
coefficients.  For each time slice, the templates are downsampled to the
appropriate sample rate.  Finally, the \SVD\ is applied to each time
slice in the sub-bank in order to produce a set of orthogonal \fir\
filters and a reconstruction matrix that maps them back to the original
templates as described in \eqref{FIXME}.  The downsampled orthogonal
\fir\ filter coefficients, the reconstruction matrix, and the time slice
boundaries are all saved to disk.

@@

We have implemented a prototype of the low latency filtering stage using an
open source signal processing environment called \gstreamer\ \cite{gstreamer}.
\gstreamer\ is a vital component of many Linux systems, providing media
playback, authoring, and streaming on devices from cell phones to desktop
computers to streaming media servers.  Given the similarities of gravitational
wave detector data to audio data it is not surprising that \gstreamer\ is
useful for our purpose. \gstreamer\ also provides some useful stock signal
processing elements such as resamplers and filters.  We have extended the
\gstreamer\ framework by developing custom elements that are available in the
\gstlal\ library~\cite{gstlal}.  Figure \ref{fig:pipeline} describes
schematically how we implement the \lloid\ algorithm using \gstlal\ and
\gstreamer\ components.
%
%
\begin{figure}[htbp]
	\includegraphics{figures/lloid-diagram.pdf}
	\caption{\label{fig:pipeline} Schematic of LLOID pipeline illustrating
signal flow.  Circles with arrows represent upsampling
\protect\includegraphics{figures/upsample-symbol.pdf} or downsampling
\protect\includegraphics{figures/downsample-symbol.pdf}.  Circles with plus
signs represent summing junctions
\protect\includegraphics{figures/adder-symbol.pdf}.  Squares
\protect\includegraphics{figures/fir-symbol.pdf} stand for FIR filters.  Sample
rate decreases from the top of the diagram to the bottom.  In this diagram each
time slice contains three \fir\ filters that are linearly combined to produce
four output channels.  In a typical pipeline the number of \fir\ filters is
much less than the number of output channels.}
\end{figure}

The filter pipeline drawn in figure \ref{fig:pipeline} consists of several
distinct stages and parallel branches.  The stages are decimation, filtering
the SVD basis, reconstruction of the original template basis, interpolation and
accumulation of the early warning and final \SNR\.  We outline the stages
below.

@@

First, the sample rate of the whitened detector data is reduced to successively
lower sample rates by decimation.  Decimation involves applying an antialiasing
filter to the data, and then downsampling by deleting samples.  We use a
192-tap \fir\ decimator provided by {\tt Gstreamer's audioresample}
element.  The detector data is provided at every power-of-two sample rate
required by the template time slices described in \eqref{FIXME}.  These
parallel decimated data streams are fed into parallel fir filter banks in the
next stage.

\subsubsection{Orthogonal \fir\ filters}

@@

The orthogonal filter output provided by the SVD must be used in linear
combinations in order to obtain the physical filter output for the template
waveforms.  This involves a matrix multiplication.  We use a \gstlal\ element
called {\tt lal\_matrixmixer} on the output of each orthogonal filter bank in
order to produce the filter output for the original waveforms in each time
slice.  At this point the time slice filter outputs are not yet at the same
sample rate, however, before they are upsampled we sum the reconstructed output
of any time slices that have a common sample rate.

@@

In the next section we present an end-to-end test of the \lloid\ algorithm
implemented in \gstlal\.

### 2011-06-02 inspiral_svd.tex
better title
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@802 ccdd05d5-56c1-461a-bf98-85f6876523b4
more honest title
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@807 ccdd05d5-56c1-461a-bf98-85f6876523b4
another tweak to the title
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@808 ccdd05d5-56c1-461a-bf98-85f6876523b4
more revisions
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@833 ccdd05d5-56c1-461a-bf98-85f6876523b4

\title[LLOID]{Improving the prospects for multimessenger astronomy with \earlywarning\ detection of compact binary coalescence.}

@@

\input{analysis}

@@

\ack{\LIGO\ was constructed by the California Institute of Technology and
Massachusetts Institute of Technology with funding from the National Science
Foundation and operates under cooperative agreement
\textsc{phy}-\oldstylenums{0107417}. CH would like to thank Ilya Mandel for
many discussions about rate estimates and the prospects of early detection.
NF would like to thank Alessandra Corsi for illuminating discussions on
astronomical motivations.

>>>>

\title[LLOID]{Technique for \earlywarning\ detection of compact binary coalescence and its implication for multimessenger astronomy}

@@

%\input{analysis}

@@

\ack{\LIGO\ was constructed by the California Institute of Technology and
Massachusetts Institute of Technology with funding from the National Science
Foundation and operates under cooperative agreement
\textsc{phy}-\oldstylenums{0107417}. CH would like to thank Ilya Mandel for
many discussions about rate estimates and the prospects of early detection.  CH
would also like to thank Patrick Brady for countless fruitful conversations
about low latency analysis methods.  NF would like to thank Alessandra Corsi
for illuminating discussions on astronomical motivations.

### 2011-06-02 results.tex
Wrapped text
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@832 ccdd05d5-56c1-461a-bf98-85f6876523b4
more revisions
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@833 ccdd05d5-56c1-461a-bf98-85f6876523b4

In this section we test our implementation of the \lloid\ method using the
pipeline described in the previous section.  We answer the following questions
%
I) What is the measured loss of \SNR\ due to the approximations used in the
\lloid\ algorithm in a realistic analysis pipeline?
%
II) What is the measured latency of the implementation described in section
\ref{sec:implementation}?
%
III) What is the measured computational cost of the implementation described in
section \ref{sec:implementation}?

@@

We expect two contributions to the \SNR\ loss to arise in our implementation of
the \lloid\ algorithm.  The first is the \SNR\ loss due to the truncation of
the SVD basis, which is fundamental, and estimates for it
exist~\cite{Cannon:2010p10398}.  The second comes from non ideal
implementations of resampling that cause signal loss.  We have measured the
effect of the resampling in the pipeline alone on a single waveform to gauge
the magnitude of the loss.  It is shown in figure \ref{fig:resamp_mm} as a
function of the quality of the {\tt audioresample} element (which is
proportional to the filter length used).  In both cases we have to compare the
\SNR\ loss to the expected \SNR\ loss that arises from the discreteness of the
template bank which is typically $\sim 3\%$.  We will consider a comparable
\SNR\ loss to be tolerable, but ideally would like to have a factor of 10 lower
loss from the \lloid\ implementation (i.e. no more than $\sim 0.3 \%$).  
%
\begin{figure}
	\label{fig:hist-interpolate}
	\begin{center}
		\includegraphics{bw_resample.pdf}
		\caption{Box-and-whisker plot of mismatch between nominal template bank and \lloid\ measured impulse respones.  The length $N^\shortuparrow$ of the interpolation filter is varied from 8 to 192.  The \textsc{svd} tolerance is kept fixed at $(1-10^{-6})$.  The lines in the center of the boxes denote the medians.  The upper and lower boundaries of the boxes show the upper and lower quartiles.  The whiskers denote the minimum and maximum mismatch over all templates.}
	\end{center}
\end{figure}

\begin{figure}
	\label{fig:hist-svd-tolerance}
	\begin{center}
		\includegraphics{bw.pdf}
		\caption{Box-and-whisker plot of mismatch between nominal
template bank and \lloid\ measured impulse respones.  The \textsc{svd}
tolerance is varied from $\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$.
The diminishing improvement is caused by hitting the resampling match limit.}
	\end{center}
\end{figure}

>>>>

In this section we test our implementation of the \lloid\ method using the
pipeline described in the previous section.  We answer the following questions:
%
I) What is the measured loss of \SNR\ due to the approximations used in the
\lloid\ algorithm in a realistic analysis pipeline?
%
II) What is the measured latency of the implementation described in section
\ref{sec:implementation}?
%
III) What is the measured computational cost of the implementation described in
section \ref{sec:implementation}?

@@

We expect two contributions to the \SNR\ loss to arise in our implementation of
the \lloid\ algorithm.  The first is the \SNR\ loss due to the truncation of
the \SVD\ basis, which is fundamental, and estimates for it
exist~\cite{Cannon:2010p10398}.  The second comes from non ideal
implementations of resampling that cause signal loss.  We have measured the
effect of the resampling in the pipeline alone on a single waveform to gauge
the magnitude of the loss.  It is shown in figure \ref{fig:resamp_mm} as a
function of the quality of the {\tt audioresample} element (which is
proportional to the filter length used).  In both cases we have to compare the
\SNR\ loss to the expected \SNR\ loss that arises from the discreteness of the
template bank which is typically $\sim 3\%$.  We will consider a comparable
\SNR\ loss to be tolerable, but ideally would like to have a factor of 10 lower
loss from the \lloid\ implementation (i.e. no more than $\sim 0.3 \%$).

It is also within reason that some \SNR loss could arise from other suboptimal
implementations of the \lloid\ algorithm in our test pipeline.  To gauge
accurately how well the pipeline is performing we tested the response to a
delta function.  This should produce time reversed versions of the original
template waveforms. By taking the innerproduct of the impulse response for each
channel with the time reversed template, we can gauge very accurately the loss
in expected \SNR\ due to the approximations we have made and any inadequacies
in our implementation.

\subsubsection{Checking the effect of the \SVD\ tolerance}

In this section we check the effect of the \SVD\ tolerance parameter defined in
~\cite{Cannon:2010p10398}.  By changing the tolerance of the reconstruction of
the physical waveforms one effects directly the mismatch that results from the
truncation of the orthogonal filter matrix.  The conditions presented here are
more complicated than in the original work~\cite{Cannon:2010p10398} due to the
inclusion of the time sliced templates and resampling.  \FIXME{However, we find
that the results are quite as expected.}  Figure \ref{fig:hist-svd-tolerance}
demonstrates that it is possible to acheive typical \SNR\ losses of $\ll1\%$.
However, we do notice that the mismatch saturates with respect to the tolerance.
This could be the effect of the resampling, or another \SNR\ loss that we did
not model or expect.  However the saturation is still an order of magnitude below our target mismatch of $\sim 0.3 \%$.  We find that an \SVD\ tolerance of
$1-1\times10^{-4}$ is adequate. 
%
\begin{figure}
	\begin{center}
		\includegraphics{bw.pdf}
		\caption{\label{fig:hist-svd-tolerance}
Box-and-whisker plot of mismatch between nominal
template bank and \lloid\ measured impulse respones.  The \textsc{svd}
tolerance is varied from $\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$.
The diminishing improvement is caused by some other limitation in the pipeline.}
	\end{center}
\end{figure}

\subsubsection{Checking the effect of the interpolation filter length}

Next, keeping the \SVD\ tolerance fixed at $1-1\times10^{-6}$, we checked the
impact of changing the filter length used to interpolate the decimated data
back up to the full sample rate.  The results are in figure
\ref{fig:hist-interpolate}.  We find that a filter length of 16 is sufficient
to produce mismatches at our target of $\sim 0.3 \%$.
%
\begin{figure}
	\begin{center}
		\includegraphics{bw_resample.pdf}
		\caption{\label{fig:hist-interpolate} Box-and-whisker plot of
mismatch between nominal template bank and \lloid\ measured impulse respones.
The length $N^\uparrow$ of the interpolation filter is varied from 8 to 192.
The \textsc{svd} tolerance is kept fixed at $(1-10^{-6})$.  The lines in the
center of the boxes denote the medians.  The upper and lower boundaries of the
boxes show the upper and lower quartiles.  The whiskers denote the minimum and
maximum mismatch over all templates.}
	\end{center}
\end{figure}
